---
title: "Arrival delays. Assignment 5"
author: "Maxim Arisov"
date: "2026-10-04"
output:
  pdf_document: default
  html_document: default
---

Approach

First, I will create a .CSV file that includes all the information mentioned in
the chart, with arrival delays provided in the instructions for Assignment 5A.
Second, I will read the information from my .CSV file into R. Then, I will use 
the tidyr and dplyr packages, which are main components of the tidyverse. I will
use dplyr to filter, select, mutate, summarize, and arrange my data, while tidyr
will be used to change the layout of my dataset by following commands: 
pivot_wider, pivot_longer, etc. In other words, the two packages named above 
will be used to tidy and transform my data.
Moreover, I will perform an analysis to compare the arrival delays for the 
two airlines.
Lastly, I will provide R code in the R Markdown file and will publish it to 
rpubs.com. Furthermore, the conclusion and narrative description of data 
manipulation and analysis will also be provided.
I will upload the Rmd file to my GitHub repository, and the URLs for 
the .Rmd file and rpubs.com web page will be provided.

GitHub links:

https://raw.githubusercontent.com/Maximus2485/Assignment-5A.-Airline-Delays/refs/heads/main/Arrival%20delays%20Assignment%205.%20.csv

Code Base  
#Lets upload the raw data in R as csv.file from GitHub.. At first I created the 
#data in Excel sheet and then downloaded to my computer as csv file. 
#Then, I uploaded it to gitHub. 

```{r}
arrival_delays <- read.csv("https://raw.githubusercontent.com/Maximus2485/Assignment-5A.-Airline-Delays/refs/heads/main/Arrival%20delays%20Assignment%205.%20.csv")

head(arrival_delays)
```
```{r}
#Lets add column name to the dataset
names(arrival_delays) <- c(
  "Airline",
  "Status",
  "Los_Angeles",
  "Phoenix",
  "San_Diego",
  "San_Francisco",
  "Seatte"
  )

arrival_delays
```
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```



```{r}
head(arrival_delays)
summary(arrival_delays)
```


```{r}
#change the character variables to numeric.Lets change the character for integer
# for Phoenix flight column
arrival_delays_num <- arrival_delays %>%
  mutate(
    across(3:7,
           ~as.numeric(gsub(',',
"",.))
  )
)
str(arrival_delays_num)

```

```{r}
# Lets check if the data has N/A
sum(is.na(arrival_delays))
```

Lets remove the row with N/A from the dataset.

```{r}
arrival_delays <- arrival_delays|>
  dplyr::ungroup()|>
  dplyr::slice(-3)

print(arrival_delays)


```
Lets fill missing airlines names
```{r}
arrival_delays_num <- arrival_delays_num %>%
  mutate(Airline=na_if(Airline, ""))%>%
  fill(Airline)

arrival_delays_num
```

#Lets change wide data to long data using tidyr
```{r}
arrival_long <- arrival_delays_num %>%
  tidyr::pivot_longer(
    cols=3:7,
    names_to = "City",
    values_to = "Flights"
  )
arrival_long
```

Lets calculate overall delay percentage
```{r}
overall <- arrival_long %>%
  group_by(Airline) %>%
  summarize(
    Total_Flights= sum(Flights),
    Delayed_Flights = sum(Flights[Status=="delayed"]),
    Delay_Percent = Delayed_Flights/Total_Flights * 100
  )
overall
#Thus, ALASKA airline had delayed flights as 501, and total flights 3775.Therefore,
# its overall delay rate was 13.27%. 
# In its turn, AM WEST airlines had 7225 flights in total where 787 flights were 
# delayed. Thus, its overall delay rate was 10.89% approximately. Therefore, 
#ALASKA while had lower number of delayed flights produced a higher percentage 
# of delayed flights comparing to AM WEST which had more delayed flights but 
# fewer delay percent. 

```

Lets calculate delay percentage by city
```{r}
by_city <- arrival_long %>%
  group_by(Airline, City)%>%
  summarize(Total_Flights= sum(Flights),
    Delayed_Flights = sum(Flights[Status=="delayed"]),
    Delay_Percent = Delayed_Flights/Total_Flights * 100
    )%>%
  ungroup()
by_city
  
```

Lets create a graph
```{r}
ggplot(by_city, mapping=aes(x=City, y=Delay_Percent, fill=Airline)) +
  geom_col(position="dodge")+
  labs(title="Flight Delay Percentage by City",
       x="City",
       y="Delay Percentage",
       fill="Airline"
       )
#Thus, we can conclude that AM WEST has highest delays in cities such as San
#Francisco and Seattle comparing to other cities. However, the lowest delays 
# percentage of AM WEST as well as ALASKA airlines noticed in the Phoenix. Also,
# AM WEST airlines have highest delay percentage in all 5 cities comparing to the 
# ALASKA ones. In its turn, ALASKA's highest percentage of delays is in the city
# of San Francisco. 
```

Lets create overall comparison graph

```{r}
ggplot(overall, mapping=aes(x=Airline, y=Delay_Percent, fill=Airline)) +
  geom_col() +
  labs(
    title="Overall Flight Delay Percentage",
    x= "Airline",
    y="Delay Percentage"
  )
```
Conclusion

The analysis demonstrated that AM WEST has highest delay percentage in all 5 
cities. Nevertheless, ALASKA has a higher overall delay percentage when all 
cities are combined. For instance, ALASKA has 13.27% of delays overall comparing
to the AM WEST with 10.89%. 
This assignment analysis and results show that both data about delay percentages
should be obtained from 2 airlines before making any conclusions; overall 
percentages and by city delay percentages should be obtained for both companies 
before making a conclusion about flight delays. 



